### 2021-07-03 ch1intro.tex
made some progress on the nightmare citations.

Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes uplifted by tectonics.
Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands, grinding boulders to sand in a machine of abrasion.
Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.

In this context river geomorphology as a scientific discipline has shifted toward more quantitative methods that enable concrete predictions about the natural world \citep{Church2005}.
Sediment transport in river channels is especially amenable to this quantitative approach since it is mechanistically the result of fluid and granular physics.
Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}. In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.

Bedload transport exerts unique influence over channel morphology and stability \citep{Church2006,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Hassan2008, Eaton2020}.
A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
Given that this problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.

Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}. The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018}.
Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values.

The approach taken in this thesis is to consider bedload transport as an aggregate result of many individual transported grains whose movements have probabilistic characteristics.
This departs from the traditional strategy of correlating transport rates to the average flow and sedimentary characteristics \citep{MeyerPeter1948, Parker1990, Wilcock2001}.
Because the trajectories of individual grains are governed by Newtonian mechanics, this approach brings reliable tools to the problem, but it is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time due to fluid turbulence and the erratic interactions of moving particles with the bed.

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and the body of earlier literature which inspired my approach.
To provide context and motivation, the thesis begins by reviewing this literature, rephrasing earlier work as necessary to indicate common themes and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport theory is to predict the downstream movement of an individual particle.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Furbish1997}.
Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the outcomes of these encounters are variable \citep{Sekine1992}.

Entrainment and deposition provide an additional layer of complexity.
Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and cause their deposition \citep{Charru2004}.
These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2012,Celik2014}. 
As a result, particles generally alternate through sequences of entrainment and deposition, but determining precisely when these events occur is impractical.

Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}, 
These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Church1994,Ferguson2002}.

@@

\subsection{Motivation: tracers and basic understanding}
An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenberger2012}. This obscures the mapping between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.

@@

Einstein's assumptions mean that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
\be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
where $\dot{x} = dx/dt$, and $\mu(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
A particular realization of this pulsed noise can be written
\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
where $N(t)$ is the number of particle entrainments in time $t$, distributed as a Poisson distribution $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1956}, but with random driving term (equation \ref{eq:einrando}).
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
	\caption{Panel (a) indicates the representation of Einstein's model as an idealized ``white shot noise", as indicated in eq. \ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$). }
	\label{fig:einsteinfig}
\end{figure}

The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1984,Moss1989}. Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
\be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1986a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
\be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/\ell}\theta(x) \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big) \ee
Here, $\mathcal{I}$ is a modified Bessel function. 
The probability distribution eq. \ref{eq:eindist} fully characterizes the dynamics of an individual particle alternating through steps and rests.

This distribution displays both advective and diffusive characteristics.
One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
\be \langle x \rangle (t) = k_E \ell t, \ee
so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$, given by the entrainment rate times the mean step length.
The rate at which one particle spreads out from another due to differences in their trajectories can be represented by the variance, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$. This gives
\be \sigma_x^2(t) = 2 k_E \ell^2 t, \ee
so particles in Einstein's model spread apart as a normal diffusion process $\sigma_x^2 = 2 D_\text{eff} t$ \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$

@@

Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles become embedded in sedimentary deposits (geomorphic range).
The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}. In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
Einstein's infinite movement velocity assumption excludes the timescales over which these processes occur.

@@

For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
\be \ee
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase. The variance, however, is different.
Computing $\sigma_x^2$ provides
\be \sigma_x^2(t) = ,\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2014}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, since fluctuations of the velocity between entrainment and deposition have been neglected.

@@

Some authors have attempted to model local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically. Early efforts impart particles with time-averaged fluid forces, typically linked to a logarithmic flow velocity profile \citep{VanRijn1984}, and some include simplified collision forces that modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992}.
Later on, researchers began to include granular interactions among particles to model collisions using the discrete element method \citep{Cundall1979, Haff1993}.
The early works utilizing this approach used a two dimensional domain with a highly simplified flow model \citep{Jiang1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow. The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, using particles as the boundary condition for the fluid \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowincklel2014,Elghannay2017,Yousefi2020}.
A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2011, Azema2012, Wachs2019}.
These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are desired for further insight into the problem.

@@

\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities with a mechanistic model including a noisy driving term to represent fluid turbulence.
They wrote, for the streamwise particle velocity $u$, a Langevin equation
\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of particle velocity fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}. The fluid drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions. The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1984,VanKampen2007} 
\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by equation \ref{eq:fanlangevin} is
\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
This is the (two-sided) exponential distribution observed in one subset of experiments.

@@

Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$, but this approach has been revised by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}. Some of these authors have argued that instead, one should calculate $P_n$ as an exceedance probability.
If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution \citep{Cox1965}:
\be p_n = \frac{(k_E T)^n}{n!}e^{-k_E T}.\ee
This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^n}{n!}e^{-k_E T}, $
so Einstein's mean sediment flux (eq. \ref{eq:einflux}) becomes
\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
Noticing that $\rho k_E$ is the entrainment rate of a single grain multiplied by the density of grains available for entrainment on the bed, we can summarize the Einstein theory as 
\be q = E \ell, \ee
where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit bed area \citep{Wilcock1997,Furbish2012}.

>>>>

Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes generated by tectonics.
Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands, grinding boulders to sand in a machine of abrasion.
Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.

In this context river geomorphology as a scientific discipline has shifted toward more quantitative methods that enable concrete predictions about the natural world \citep{Church2005,Church2010}.
Sediment transport in river channels is especially amenable to this quantitative approach since it is mechanistically the result of fluid and granular physics.
Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947,Turowski2020}. In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.

Bedload transport exerts unique influence over channel morphology and stability \citep{Church2006,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Hassan2008, Eaton2020}.
A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement, since this relates to morphological change.
Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
Given that this problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.

Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}. The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018, Masteller2019, Pretzlav2020}.
Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values \citep{Ancey2020b}.

The approach taken in this thesis is to consider bedload transport as an aggregate result of many individual transported grains whose movements have probabilistic characteristics.
This departs from the traditional strategy of correlating transport rates to the average flow and sedimentary characteristics \citep{MeyerPeter1948, Parker1990, Wilcock2001}.
Because the trajectories of individual grains are governed by Newtonian mechanics, this approach brings powerful tools to the problem, but it is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time due to fluid turbulence and the erratic interactions of moving particles with the bed.

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and drawing upon a wide interdisciplinary literature.
To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport theory is to predict the downstream movement of an individual particle.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Coleman1967,Schmeeckle2007}.
Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Martin2013}.

Entrainment and deposition provide an additional layer of complexity.
Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and induce their deposition \citep{Charru2004}.
These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2010,Celik2014}. 
As a result, particles generally alternate through sequences of entrainment and deposition, but the exact times at which these alternations occur is impossible to characterize.

Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}, 
These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Hassan1994,Ferguson2002}.

@@

\subsection{Motivation: tracers and basic understanding}
An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013}. This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.

@@

Einstein's assumptions mean that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
\be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
where $\dot{x} = dx/dt$, and $\mu(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
A particular realization of this pulsed noise can be written
\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
where $N(t)$ is the number of particle entrainments in time $t$, distributed as a Poisson distribution $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein2001}, but with random driving term (equation \ref{eq:einrando}).
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
	\caption{Panel (a) indicates the representation of Einstein's model as an idealized ``white shot noise", as indicated in eq. \ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$). }
	\label{fig:einsteinfig}
\end{figure}

The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1989,Moss1989}. Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
\be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1992a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
\be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/\ell}\theta(x) \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big) \ee
Here, $\mathcal{I}$ is a modified Bessel function. 
The probability distribution eq. \ref{eq:eindist} fully characterizes the dynamics of an individual particle alternating through steps and rests.

This distribution displays both advective and diffusive characteristics.
One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
\be \langle x \rangle (t) = k_E \ell t, \ee
so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$, given by the entrainment rate times the mean step length.
The rate at which one particle spreads out from another due to differences in their trajectories can be represented by the variance, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$. This gives
\be \sigma_x^2(t) = 2 k_E \ell^2 t, \ee
so particles in Einstein's model spread apart as a normal diffusion process $\sigma_x^2 = 2 D_\text{eff} t$ \citep{Sokolov2012}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$

@@

Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles become embedded in sedimentary deposits (geomorphic range).
The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}. In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
Einstein's infinite movement velocity assumption excludes the timescales over which these processes occur.

@@

For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
\be \ee
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase. The variance, however, is different.
Computing $\sigma_x^2$ provides
\be \sigma_x^2(t) = ,\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, since fluctuations of the velocity between entrainment and deposition have been neglected.

@@

Some authors have attempted to model local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically. Early efforts impart particles with time-averaged fluid forces, typically linked to a logarithmic flow velocity profile \citep{VanRijn1984}, and some include simplified collision forces that modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992}.
Later on, researchers began to include granular interactions among particles to model collisions using the discrete element method \citep{Cundall1979a, Haff1993}.
The early works utilizing this approach used a two dimensional domain with a highly simplified flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow. The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, using particles as the boundary condition for the fluid \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2012, Azema2012, Wachs2019}.
These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are desired for further insight into the problem.

@@

\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities with a mechanistic model including a noisy driving term to represent fluid turbulence.
They wrote, for the streamwise particle velocity $u$, a Langevin equation
\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of particle velocity fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}. The fluid drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions. The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by equation \ref{eq:fanlangevin} is
\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
This is the (two-sided) exponential distribution observed in one subset of experiments.

@@

Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$, but this approach has been revised by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}. Some of these authors have argued that instead, one should calculate $P_n$ as an exceedance probability.
If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution \citep{Cox1965}:
\be p_n = \frac{(k_E T)^n}{n!}e^{-k_E T}.\ee
This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^n}{n!}e^{-k_E T}, $
so Einstein's mean sediment flux (eq. \ref{eq:einflux}) becomes
\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
Noticing that $\rho k_E$ is the entrainment rate of a single grain multiplied by the density of grains available for entrainment on the bed, we can summarize the Einstein theory as 
\be q = E \ell, \ee
where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit bed area \citep{Wilcock1997a,Furbish2012}.

Alongside this central result $q=E\ell$ for the mean flux, one of Einstein's most influential and enduring ideas was his formulation of the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
The original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
The entrainment rate is calculated from the exceedance probability of the turbulent lift over the weight, providing an alternative to the critical shear stress concept which explicitly incorporates turbulent fluctuations in the fluid flow. This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
random) resisting quantities \citep{Grass1970} is a key part of Einstein's legacy. \citet{Paintal1971} made a significant extension
by including the random supporting geometry of bed particles into the force balance. More recently, \citet{Tregnaghi2012}
ammended the theory to include both force magnitude and duration \citep{Diplas2008, Valyrakis2013, Celik2014}.
Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.
